---
title: "Introduction to Machine Learning: Zomato & Swiggy Case Study"
output:
  html_notebook:
    toc: true
    theme: flatly
    highlight: tango
---

# Introduction

Welcome to your first hands-on Machine Learning (ML) notebook! This notebook is designed for beginners with no prior ML experience. We will use a simulated dataset representing **Zomato and Swiggy** restaurant profiles to explore four foundational machine learning algorithms. 

By using R's intuitive formula syntax (`y ~ x`), we will learn how to:
1. **Predict continuous numbers** (Linear Regression)
2. **Classify binary outcomes** (Logistic Regression)
3. **Build logical decision flowcharts** (Decision Trees)
4. **Group similar items automatically** (K-Means Clustering)

---

## Step 0: Install and Load Required Packages

We will use three key R packages:
* `ggplot2` for high-quality data visualizations.
* `rpart` for building recursive partitioning decision trees.
* `rpart.plot` for rendering visual decision tree flowcharts.

```{r setup, message=FALSE, warning=FALSE}
# Install packages if you do not have them:
# install.packages("rpart")
# install.packages("rpart.plot")
# install.packages("ggplot2")

library(rpart)
library(rpart.plot)
library(ggplot2)

# Set seed for reproducibility so we all get the exact same results
set.seed(42)
```

---

## Step 1: Create the Synthetic Dataset

To ensure this notebook runs instantly without troubleshooting external file paths or missing CSVs, we will programmatically generate a dataset of 200 restaurants mimicking real-world Zomato and Swiggy features:
* `average_cost_two`: Cost for two diners (ranging from 150 to 2500 INR).
* `distance_center_km`: Distance from the city center (0.5 to 15 km).
* `rating`: Average customer rating (1.5 to 4.9 stars).
* `votes`: Number of reviews/ratings submitted by customers.

We will inject realistic business logic so our models can discover true patterns:
1. **Linear Relationship**: Revenue increases with more votes and higher food costs, but decreases with distance.
2. **Binary Relationship**: Premium/expensive restaurants are much more likely to offer online delivery.
3. **Non-linear Tiering**: Price categories are bounded by specific cost brackets.

```{r generate_data}
n_restaurants <- 200

restaurant_data <- data.frame(
  restaurant_id = 1:n_restaurants,
  platform = sample(c("Zomato", "Swiggy"), n_restaurants, replace = TRUE),
  average_cost_two = round(runif(n_restaurants, min = 150, max = 2500)),
  distance_center_km = round(runif(n_restaurants, min = 0.5, max = 15), 1),
  rating = round(runif(n_restaurants, min = 1.5, max = 4.9), 1),
  votes = round(runif(n_restaurants, min = 10, max = 1500))
)

# 1. Generate revenue (Continuous variable)
restaurant_data$estimated_revenue <- round(
  (restaurant_data$votes * 1.5) + 
  (restaurant_data$average_cost_two * 0.4) - 
  (restaurant_data$distance_center_km * 50) + 
  rnorm(n_restaurants, mean = 500, sd = 100)
)
# Force minimum revenue to be 100 INR
restaurant_data$estimated_revenue[restaurant_data$estimated_revenue < 100] <- 100

# 2. Generate online delivery status (Binary 1/0 variable using sigmoid probability)
prob_delivery <- 1 / (1 + exp(-(-1 + 0.0015 * restaurant_data$average_cost_two)))
restaurant_data$has_online_delivery <- rbinom(n_restaurants, 1, prob_delivery)

# 3. Categorize price range (Multi-class factor)
restaurant_data$price_range <- factor(cut(restaurant_data$average_cost_two, 
                                   breaks = c(0, 400, 800, 1500, Inf), 
                                   labels = c("Budget", "Moderate", "Premium", "Luxury")))

# Preview the data
head(restaurant_data)
```

---

## Section 1: Linear Regression (Predicting a Continuous Number)

### The Concept
Linear Regression finds a straight-line relationship between one or more input variables (features) and a numeric output (target). It aims to find the line that minimizes the total prediction error (residuals).

### Our Goal
Predict a restaurant's **Estimated Monthly Revenue** based on its distance from the city center and customer votes.

```{r linear_regression}
# formula syntax: target ~ feature_1 + feature_2
linear_model <- lm(estimated_revenue ~ distance_center_km + votes, data = restaurant_data)

# Review model statistics
summary(linear_model)
```

### Visualizing the Fit
Let's plot how estimated revenue scales with customer votes, color-coded by distance:

```{r plot_linear}
ggplot(restaurant_data, aes(x = votes, y = estimated_revenue, color = distance_center_km)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_smooth(method = "lm", col = "red", se = FALSE, size = 1.2) +
  labs(title = "Linear Regression: Revenue vs. Votes",
       x = "Number of Votes (Popularity)",
       y = "Estimated Monthly Revenue (INR)",
       color = "Distance (KM)") +
  theme_minimal()
```

---

## Section 2: Logistic Regression (Predicting a Binary Classification)

### The Concept
When we want to predict a binary category (Yes/No, 1/0), standard linear regression fails because it can predict numbers below 0 or above 1. **Logistic Regression** solves this by wrapping the linear equation in a **Sigmoid function**, mapping any input into a probability between 0% and 100%.

### Our Goal
Predict whether a restaurant **offers online delivery** (`has_online_delivery`) based on its average cost for two and its rating.

```{r logistic_regression}
# glm() with family = binomial specifies a logistic regression model
logistic_model <- glm(has_online_delivery ~ average_cost_two + rating, 
                      data = restaurant_data, 
                      family = binomial)

# Review model statistics
summary(logistic_model)
```

---

## Section 3: Decision Trees (Flowchart-Based Rules)

### The Concept
Instead of mathematical lines or probabilities, a **Decision Tree** splits data step-by-step using a series of logical "If-Else" rules. It selects features that divide the data into groups that are as pure (homogenous) as possible, measuring success via metrics like **Gini Impurity** or **Entropy**.

### Our Goal
Classify a restaurant's pricing tier (`price_range`: Budget, Moderate, Premium, Luxury) based on its average cost, platform, and rating.

```{r decision_tree}
# Build the classification tree
tree_model <- rpart(price_range ~ average_cost_two + rating + platform, 
                    data = restaurant_data, 
                    method = "class")

# Plot the visual decision tree flowchart
rpart.plot(tree_model, 
           type = 2, 
           extra = 104, 
           under = TRUE, 
           fallen.leaves = TRUE, 
           main = "Decision Tree: Classifying Price Range Tier")
```

---

## Section 4: K-Means Clustering (Unsupervised Grouping)

### The Concept
Unlike the previous three models (which are **Supervised Learning** because they had correct target answers to train on), **K-Means Clustering** is **Unsupervised**. It does not know any correct labels; instead, it automatically groups similar restaurants together in space based on how physically close their features are.

*Note: We must use the `scale()` function first. Because average cost is measured in thousands (150 - 2500) and ratings in single digits (1.5 - 4.9), the cost feature would mathematically overwhelm the rating feature during distance calculations if they aren't standardized to the same scale.*

### Our Goal
Group our restaurants into 3 distinct marketing segments based on their average cost and rating.

```{r kmeans_clustering}
# Scale/standardize features to give them equal mathematical weight
cluster_features <- scale(restaurant_data[, c("average_cost_two", "rating")])

# Run K-means clustering to find 3 centers
kmeans_result <- kmeans(cluster_features, centers = 3, nstart = 20)

# Append the discovered cluster assignments back to our main dataset
restaurant_data$cluster <- factor(kmeans_result$cluster)

# Let's inspect the cluster centers on original scale values
aggregate(restaurant_data[, c("average_cost_two", "rating")], 
          by = list(Cluster = restaurant_data$cluster), 
          FUN = mean)
```

### Visualizing the Discovered Clusters
Let's chart the 3 natural marketing groups identified by the algorithm:

```{r plot_clusters}
ggplot(restaurant_data, aes(x = average_cost_two, y = rating, color = cluster)) +
  geom_point(size = 3, alpha = 0.8) +
  labs(title = "K-Means Clustering: Identifying 3 Restaurant Segments",
       x = "Average Cost for Two (INR)",
       y = "Customer Rating (Stars)",
       color = "Discovered Cluster") +
  theme_minimal()
```

---

## Summary for Students

| Aspect | Supervised Models (Linear/Logistic/Trees) | Unsupervised Models (K-Means) |
| :--- | :--- | :--- |
| **Data Requirements** | Requires historical inputs **and** correct labels ($y$) | Only requires inputs; discovers structure on its own |
| **Typical Goal** | Predicting a numeric value or assigning a category | Grouping similar points together to find patterns |
| **Real-world Use Case** | Estimating food delivery time or predicting order volume | Segmenting customers into spending behavior groups |
