---
title: "BDA Session — 2026-09-29"
subtitle: "In-Database Analytics: Advanced SQL & Apache MADlib"
author: "BDA Instructor"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: darkly
    highlight: zenburn
    toc: true
    toc_float: true
    number_sections: true
---

<!--
## Where to run this notebook

This `.Rmd` works on any platform with R 4.x and the `rmarkdown` package.
Pick whichever fits your setup:

| Platform | Best for | How to open |
|----------|----------|-------------|
| **posit.cloud** *(class default)* | Zero setup, browser-only | New Project → Upload → knit |
| **Quarto + RStudio Desktop** *(recommended next step)* | Local dev, richer output formats | Rename `.Rmd` → `.qmd`; same syntax, better themes & multi-language support |
| **VS Code + R Extension** | Engineers who live in VS Code | `Ctrl+Shift+K` knits; install `rmarkdown` pkg first |
| **GitHub Codespaces** | Reproducible, version-controlled cloud | Add `devcontainer.json` with `rocker/tidyverse`; free 60 h/month |
| **Binder** | Sharing a live notebook with zero install | Push repo to GitHub → mybinder.org → paste URL → instant RStudio in browser |
| **JupyterHub + IRkernel** | Multi-user classroom or org | Instructor provisions once; students log in via browser |

> **Note:** Content synthesized from training knowledge — YouTube transcripts were unavailable
> due to proxy restrictions. Flag: `synthesized: no transcript available`
-->

```{r setup, include=FALSE}
# Global chunk options — apply to all chunks below
knitr::opts_chunk$set(
  echo    = TRUE,    # show code in output
  warning = FALSE,   # suppress package warnings (cleaner output)
  message = FALSE    # suppress messages like library load notices
)

# Fix the random seed so results are perfectly reproducible
set.seed(42)

# Core packages — available on posit.cloud with no install
library(dplyr)    # data manipulation (mirrors SQL syntax closely)
library(ggplot2)  # data visualisation
```

---

# Topic 1: Advanced SQL Window Functions & Greenplum Extensions

SQL window functions let you compute aggregates **over a set of rows related to the current row** — without collapsing the result set into a summary like `GROUP BY` does. Think of a window as a sliding frame of data that moves row by row.

In **Greenplum** (an MPP PostgreSQL database), these functions run in parallel across every segment simultaneously. For a platform tracking hundreds of millions of orders, that parallelism is the difference between an 18-minute analytical job and an overnight batch.

In this notebook, we use R and `dplyr` to *illustrate* the same logic SQL window functions implement. The equivalences are direct: wherever you see `group_by() + mutate()` in dplyr, the SQL counterpart is `PARTITION BY` inside `OVER()`.

## Synthetic dataset: restaurant weekly sales

```{r create-data}
# Build a synthetic Zomato/Swiggy-style dataset
# 5 cities × 6 restaurants × 8 weeks = 240 rows

cities      <- c("Mumbai", "Delhi", "Bangalore", "Chennai", "Hyderabad")
n_rest      <- 6    # restaurants per city
n_weeks     <- 8    # weeks of history

# Generate restaurant IDs per city
restaurant_data <- expand.grid(
  city          = cities,
  restaurant_id = 1:n_rest,   # 1–6 within each city
  week_num      = 1:n_weeks
) %>%
  mutate(
    # Unique restaurant ID across all cities
    restaurant_id = paste0(substr(city, 1, 3), "_R", restaurant_id),

    # Week-ending date starting from 2026-07-06
    week_ending   = as.Date("2026-07-06") + (week_num - 1) * 7,

    # Weekly revenue: base (city effect) + restaurant effect + noise
    base_revenue  = case_when(
      city == "Mumbai"    ~ 180000,
      city == "Delhi"     ~ 165000,
      city == "Bangalore" ~ 155000,
      city == "Chennai"   ~ 130000,
      TRUE                ~ 120000   # Hyderabad
    ),
    # Each restaurant has a persistent quality multiplier (0.6–1.4)
    quality_mult  = rep(
      runif(length(cities) * n_rest, 0.6, 1.4),
      each = n_weeks
    ),
    weekly_revenue = as.integer(base_revenue * quality_mult *
                                  rnorm(n(), mean = 1, sd = 0.08))
  ) %>%
  select(restaurant_id, city, week_ending, weekly_revenue) %>%
  arrange(city, restaurant_id, week_ending)

# Preview the first few rows
head(restaurant_data, 10)
```

## RANK() within partition — "which restaurant is #1 in Mumbai this week?"

```{r window-rank}
# SQL equivalent:
#   RANK() OVER (PARTITION BY city, week_ending ORDER BY weekly_revenue DESC)
#
# dplyr equivalent: group_by the partition columns, then mutate (not summarise!)
ranked <- restaurant_data %>%
  group_by(city, week_ending) %>%           # PARTITION BY city, week_ending
  mutate(
    city_rank = rank(-weekly_revenue,        # ORDER BY weekly_revenue DESC
                     ties.method = "min")    # ties share the lowest rank number
  ) %>%
  ungroup()

# Show the top-ranked restaurant in each city for the most recent week
ranked %>%
  filter(week_ending == max(week_ending), city_rank == 1) %>%
  select(city, restaurant_id, weekly_revenue, city_rank) %>%
  arrange(desc(weekly_revenue))
```

The key insight: `group_by() + mutate()` keeps **every row** (like a window function), while `group_by() + summarise()` collapses them (like GROUP BY). SQL's `OVER()` clause works the same way.

## LAG() — week-over-week revenue change

```{r window-lag}
# SQL equivalent:
#   weekly_revenue - LAG(weekly_revenue, 1) OVER (
#       PARTITION BY restaurant_id ORDER BY week_ending
#   ) AS wow_delta
#
# In dplyr: group by restaurant, order by date, then lag()

wow <- restaurant_data %>%
  group_by(restaurant_id) %>%               # PARTITION BY restaurant_id
  arrange(week_ending) %>%                  # ORDER BY week_ending
  mutate(
    prev_revenue = lag(weekly_revenue, 1),  # LAG(weekly_revenue, 1)
    wow_delta    = weekly_revenue - prev_revenue,   # week-over-week change
    wow_pct      = round(100 * wow_delta / prev_revenue, 1)  # as a percentage
  ) %>%
  ungroup()

# Which restaurants had the biggest single-week surge?
wow %>%
  filter(!is.na(wow_pct)) %>%
  arrange(desc(wow_pct)) %>%
  select(restaurant_id, city, week_ending, weekly_revenue, wow_pct) %>%
  head(8)
```

## Cumulative SUM and NTILE (deciles)

```{r window-cumsum-ntile}
# SQL equivalent:
#   SUM(weekly_revenue) OVER (
#       PARTITION BY restaurant_id ORDER BY week_ending
#       ROWS BETWEEN UNBOUNDED PRECEDING AND CURRENT ROW
#   ) AS cumulative_revenue
#
#   NTILE(10) OVER (PARTITION BY city, week_ending ORDER BY weekly_revenue DESC)

enriched <- ranked %>%
  group_by(restaurant_id) %>%
  arrange(week_ending) %>%
  mutate(
    cumulative_revenue = cumsum(weekly_revenue)   # running total
  ) %>%
  group_by(city, week_ending) %>%
  mutate(
    # Assign each restaurant to a decile (1 = top 10%, 10 = bottom 10%)
    revenue_decile = ntile(desc(weekly_revenue), 10)
  ) %>%
  ungroup()

# Average cumulative revenue by city after 8 weeks
enriched %>%
  filter(week_ending == max(week_ending)) %>%
  group_by(city) %>%
  summarise(avg_cumulative = mean(cumulative_revenue),
            .groups = "drop") %>%
  arrange(desc(avg_cumulative))
```

## Visualisation: revenue trend with ranking overlay

```{r plot-trend, fig.width=11, fig.height=5}
# Plot weekly revenue for the #1 restaurant in each city
top_restaurants <- ranked %>%
  group_by(city) %>%
  filter(city_rank == 1) %>%
  # pick the restaurant that was #1 most often
  count(restaurant_id) %>%
  slice_max(n, n = 1) %>%
  pull(restaurant_id)

ranked %>%
  filter(restaurant_id %in% top_restaurants) %>%
  ggplot(aes(x = week_ending, y = weekly_revenue / 1000,
             colour = city, group = restaurant_id)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2.5) +
  scale_colour_brewer(palette = "Set2") +
  labs(
    title   = "Top Restaurant Weekly Revenue by City",
    subtitle = "Each line = the restaurant ranked #1 most frequently in that city",
    x       = "Week Ending",
    y       = "Revenue (₹ thousands)",
    colour  = "City"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```

The chart mirrors what `RANK() OVER (PARTITION BY city...)` returns in Greenplum — each city's leader tracked independently across time.

---

# Topic 2: Apache MADlib — In-Database Machine Learning

Apache MADlib is a library of ML algorithms implemented as SQL user-defined functions. You call `madlib.logregr_train(...)` the same way you call `SUM()` or `RANK()`. The difference: MADlib's functions run on every Greenplum segment in parallel, so training a logistic-regression model on 50 million rows takes minutes instead of hours.

**The core philosophy:** bring the algorithm to the data, not the data to the algorithm.

Since MADlib runs inside PostgreSQL/Greenplum (not in R), this notebook *simulates* the same computations using base R functions (`lm()`, `glm()`, `kmeans()`). Wherever you see a comment showing the MADlib SQL call, that's exactly what you would execute in production.

## Synthetic customer dataset for churn prediction

```{r customer-data}
# Simulate a customer behaviour table
# In production this would be a Greenplum table: customer_features
n_customers <- 2000   # illustrative sample

customers <- data.frame(
  customer_id    = paste0("C", sprintf("%05d", 1:n_customers)),

  # Feature 1: months on platform (0–36)
  tenure_months  = sample(1:36, n_customers, replace = TRUE),

  # Feature 2: average monthly spend in ₹
  monthly_spend  = round(rnorm(n_customers, mean = 650, sd = 250)),

  # Feature 3: support tickets in last 6 months (frustration signal)
  support_tickets = rpois(n_customers, lambda = 1.2),

  # Feature 4: login frequency in last 30 days (engagement signal)
  login_freq_30d  = rpois(n_customers, lambda = 8)
) %>%
  mutate(
    monthly_spend = pmax(monthly_spend, 50),    # no negative spend

    # Churn label: high tickets + low engagement + short tenure → more likely to churn
    churn_prob = plogis(                        # logistic function → probability
      -2.5
      + 0.4  * support_tickets
      - 0.04 * login_freq_30d
      - 0.05 * tenure_months
      - 0.001 * monthly_spend
    ),
    churned = rbinom(n_customers, 1, churn_prob)   # 0 = retained, 1 = churned
  ) %>%
  select(-churn_prob)

cat("Churn rate in sample:", round(100 * mean(customers$churned), 1), "%\n")
head(customers, 5)
```

## Train / test split

```{r train-test-split}
# 70/30 split — in Greenplum you'd use a random() column filter in your SQL
train_idx  <- sample(nrow(customers), 0.7 * nrow(customers))
train_data <- customers[train_idx, ]
test_data  <- customers[-train_idx, ]

cat("Training rows:", nrow(train_data), "\n")
cat("Test rows    :", nrow(test_data),  "\n")
```

## MADlib analogue: logistic regression with glm()

```{r logistic-regression}
# ------------------------------------------------------------------
# MADlib SQL (what you'd run in Greenplum):
#
#   SELECT madlib.logregr_train(
#       'customer_features',
#       'churn_model',
#       'churned',
#       'ARRAY[tenure_months, monthly_spend, support_tickets, login_freq_30d]',
#       NULL, 20, 'irls'
#   );
#
# R equivalent using glm() with family = binomial (logistic):
# ------------------------------------------------------------------

churn_model <- glm(
  churned ~ tenure_months + monthly_spend + support_tickets + login_freq_30d,
  data   = train_data,
  family = binomial(link = "logit")   # logistic regression
)

# Inspect coefficients (like querying the churn_model table in MADlib)
summary_tbl <- data.frame(
  feature     = names(coef(churn_model)),
  coefficient = round(coef(churn_model), 4),
  p_value     = round(summary(churn_model)$coefficients[, 4], 4)
)
summary_tbl
```

Positive coefficients increase churn probability; negative ones reduce it. `support_tickets` should be the strongest positive predictor — we built that signal into the data.

## Score customers and evaluate

```{r scoring-and-eval}
# ------------------------------------------------------------------
# MADlib SQL (scoring):
#
#   SELECT customer_id,
#          madlib.logregr_predict(coef,
#              ARRAY[tenure_months, monthly_spend,
#                    support_tickets, login_freq_30d]) AS predicted_churn,
#          madlib.logregr_predict_prob(coef, ...) AS churn_probability
#   FROM   customer_features JOIN churn_model ON true
#   ORDER  BY churn_probability DESC;
#
# R equivalent:
# ------------------------------------------------------------------

test_data <- test_data %>%
  mutate(
    # Predicted probability of churning
    churn_probability = predict(churn_model,
                                newdata = test_data,
                                type    = "response"),

    # Classify: threshold at 0.5
    predicted_churn   = ifelse(churn_probability >= 0.5, 1, 0)
  )

# Confusion matrix
conf_matrix <- table(
  Predicted = test_data$predicted_churn,
  Actual    = test_data$churned
)
print(conf_matrix)

# Accuracy
accuracy <- mean(test_data$predicted_churn == test_data$churned)
cat("\nAccuracy:", round(100 * accuracy, 1), "%\n")
```

## Visualisation: churn probability distribution

```{r churn-plot, fig.width=9, fig.height=4}
ggplot(test_data, aes(x = churn_probability,
                      fill = factor(churned, labels = c("Retained", "Churned")))) +
  geom_histogram(binwidth = 0.05, position = "identity",
                 alpha = 0.65, colour = "white") +
  scale_fill_manual(values = c("Retained" = "#2ecc71", "Churned" = "#e74c3c")) +
  labs(
    title    = "Predicted Churn Probability Distribution",
    subtitle = "MADlib logregr_predict_prob() equivalent — higher = higher churn risk",
    x        = "Predicted Probability of Churning",
    y        = "Number of Customers",
    fill     = "Actual Label"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```

## MADlib analogue: k-means customer segmentation

```{r kmeans-segmentation}
# ------------------------------------------------------------------
# MADlib SQL:
#   SELECT madlib.kmeans(
#       'customer_features',
#       'ARRAY[monthly_spend, login_freq_30d, tenure_months]',
#       4,                  -- k = 4 clusters
#       'madlib.dist_norm2' -- Euclidean distance
#   );
#
# R equivalent:
# ------------------------------------------------------------------

# Scale features before clustering (important! different units)
features_scaled <- scale(customers[, c("monthly_spend",
                                       "login_freq_30d",
                                       "tenure_months")])

kmeans_result <- kmeans(
  features_scaled,
  centers  = 4,     # k = 4 segments
  nstart   = 25,    # try 25 random starts, keep best
  iter.max = 100    # max iterations per start
)

# Attach cluster labels back to customer data
customers$segment <- factor(kmeans_result$cluster,
                             labels = paste0("Segment ", 1:4))

# Profile each segment
customers %>%
  group_by(segment) %>%
  summarise(
    n_customers    = n(),
    avg_spend      = round(mean(monthly_spend)),
    avg_logins     = round(mean(login_freq_30d), 1),
    avg_tenure     = round(mean(tenure_months), 1),
    churn_rate_pct = round(100 * mean(churned), 1),
    .groups = "drop"
  ) %>%
  arrange(desc(avg_spend))
```

```{r kmeans-plot, fig.width=9, fig.height=5}
ggplot(customers, aes(x = monthly_spend, y = login_freq_30d,
                      colour = segment, shape = factor(churned))) +
  geom_point(alpha = 0.55, size = 2) +
  scale_colour_brewer(palette = "Set1") +
  scale_shape_manual(values = c(16, 4),
                     labels = c("Retained", "Churned")) +
  labs(
    title    = "Customer Segments — k-Means (k=4)",
    subtitle = "MADlib kmeans() equivalent  ·  shape = churn outcome",
    x        = "Monthly Spend (₹)",
    y        = "Login Frequency (30-day)",
    colour   = "Segment",
    shape    = "Outcome"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```

Each cluster represents a distinct customer behaviour pattern. In production, MADlib's `kmeans()` would process millions of customers in parallel across Greenplum segments, with results written directly into a scoring table — no data movement required.

---

# References

> ⚠ Content synthesized from training knowledge — YouTube transcripts unavailable (proxy blocked).

## Videos (watch independently)

- [Mastering Window Functions in SQL for Data Analysis: 1 Hour Mini-Course](https://youtube.com/watch?v=JpwwbwGSll4)
- [The Ultimate SQL Window Functions Course](https://youtube.com/watch?v=Wvg4PjbMTO8)
- [Greenplum SQL Scripts for Data Analysts](https://youtube.com/watch?v=2C5Vx4TuH1g)
- [Greenplum Database — Massively Parallel PostgreSQL for Analytics](https://youtube.com/watch?v=X8MmLeFZ1Os)
- [Machine Learning on Greenplum with Apache MADlib](https://youtube.com/watch?v=1yDn4H4zxtk)
- [Classification Models in Greenplum using Apache MADlib](https://youtube.com/watch?v=ht9y50oZUEU)
- [Parallel Postgres: Gaming MAD skills with Apache MADlib and Greenplum](https://youtube.com/watch?v=qgQy_AnswNk)

## Official Documentation

- [Apache MADlib Documentation](https://madlib.apache.org/docs/latest/)
- [Greenplum Analytics Functions Reference](https://docs.vmware.com/en/VMware-Greenplum/)
- [PostgreSQL Window Functions Tutorial](https://www.postgresql.org/docs/current/tutorial-window.html)
- [dplyr Window Functions Vignette](https://dplyr.tidyverse.org/articles/window-functions.html)
