---
title: "Hive & Pig: Querying and Cleaning Unstructured Data at Scale"
subtitle: "BDA POP — Session 09/22/2026"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Why this session exists

Last week's session established that Hadoop's HDFS+YARN pair removes the single-machine ceiling, and
that MapReduce is the programming model that runs on top of it — but it also named Hive and Pig only
in passing, as tools that "compile down to MapReduce so nobody writes it by hand." That's true, but it
skips the part that actually matters day to day: what does an analyst asking a question, or an
engineer cleaning yesterday's raw logs, actually type? This week answers that directly. Hive gives a
SQL-shaped front end over data already sitting in HDFS — the same partition-pruning and table-design
choices a database administrator makes, applied to a cluster instead of one server. Pig gives a
scripting front end for the messier job that has to happen first: turning raw, half-broken order and
delivery logs into the clean, structured rows Hive can query in the first place. Both exist because
"write a Map function and a Reduce function" is the wrong level of abstraction for either job.

### STEP 0: LIBRARIES

```{r libraries}
# Base R only this week -- everything below (split, aggregate, table) ships
# with R itself, no install.packages() needed on posit.cloud.
```

### STEP 1: THE RUNNING DATASET, EXTENDED TO ORDER-LEVEL LOGS

Same self-generated Zomato/Swiggy restaurant dataset as previous sessions — `set.seed(42)`, no
external files. This week adds two things a Hive/Pig session actually needs: a `city` column (so we
have something realistic to partition by) and an `orders` table generated *from* the restaurants —
many order-log rows per restaurant, the way a real platform's raw event stream would arrive, with a
deliberate slice of dirty rows mixed in.

```{r dataset}
set.seed(42)
n_restaurants <- 200

restaurant_data <- data.frame(
  restaurant_id = 1:n_restaurants,
  platform = sample(c("Zomato", "Swiggy"), n_restaurants, replace = TRUE),
  average_cost_two = round(runif(n_restaurants, min = 150, max = 2500)),
  distance_center_km = round(runif(n_restaurants, min = 0.5, max = 15), 1),
  rating = round(runif(n_restaurants, min = 1.5, max = 4.9), 1),
  votes = round(runif(n_restaurants, min = 10, max = 1500))
)

# More votes + higher cost => higher estimated revenue (linear relationship)
restaurant_data$estimated_revenue <- round(
  (restaurant_data$votes * 1.5) + (restaurant_data$average_cost_two * 0.4) -
    (restaurant_data$distance_center_km * 50) + rnorm(n_restaurants, mean = 500, sd = 100)
)
restaurant_data$estimated_revenue[restaurant_data$estimated_revenue < 100] <- 100

# Expensive restaurants are more likely to offer online delivery (binary relationship)
prob_delivery <- 1 / (1 + exp(-(-1 + 0.0015 * restaurant_data$average_cost_two)))
restaurant_data$has_online_delivery <- rbinom(n_restaurants, 1, prob_delivery)

# Price tier from cost (non-linear rule relationship)
restaurant_data$price_range <- factor(cut(restaurant_data$average_cost_two,
                                   breaks = c(0, 400, 800, 1500, Inf),
                                   labels = c("Budget", "Moderate", "Premium", "Luxury")))

# --- EXTENSION 1: a city per restaurant -- this is the column we'll PARTITION BY ---
cities <- c("Bangalore", "Mumbai", "Delhi", "Hyderabad", "Chennai", "Pune")
restaurant_data$city <- sample(cities, n_restaurants, replace = TRUE)

# --- EXTENSION 2: order-level logs, many rows per restaurant, spanning 14 days ---
# Real platforms don't hand Hive one row per restaurant -- they hand it millions of
# raw order-event rows. We simulate a small slice of that: 15-40 orders per
# restaurant, each with its own date and delivery time, plus a deliberately dirty
# slice (~4%) with a missing or negative delivery time -- exactly the kind of
# garbage a raw ingestion pipeline produces before anyone cleans it.
order_id <- 1L
order_rows <- list()
for (r in 1:n_restaurants) {
  n_orders <- sample(15:40, 1)
  order_dates <- as.character(Sys.Date() - sample(0:13, n_orders, replace = TRUE))
  base_time <- 20 + restaurant_data$distance_center_km[r] * 2
  delivery_time <- round(rnorm(n_orders, mean = base_time, sd = 6))
  order_rows[[r]] <- data.frame(
    order_id = order_id:(order_id + n_orders - 1),
    restaurant_id = r,
    city = restaurant_data$city[r],
    order_date = order_dates,
    delivery_time_minutes = delivery_time
  )
  order_id <- order_id + n_orders
}
orders <- do.call(rbind, order_rows)

# Inject dirty rows: ~4% of raw log rows get a missing or nonsensical delivery time.
# This is what "unstructured/raw log data" actually looks like before a Pig-style
# cleaning pass -- not free text, just untrustworthy fields mixed into the stream.
n_orders_total <- nrow(orders)
dirty_idx <- sample(seq_len(n_orders_total), size = round(0.04 * n_orders_total))
half <- length(dirty_idx) %/% 2
orders$delivery_time_minutes[dirty_idx[seq_len(half)]] <- NA
orders$delivery_time_minutes[dirty_idx[(half + 1):length(dirty_idx)]] <- -1

cat("Restaurants:", n_restaurants, " | Order-log rows:", n_orders_total,
    " | Cities:", length(cities), " | Date span: 14 days\n")
head(orders, 5)
```

---

## Hive: a SQL front end over data that never leaves HDFS

Hive doesn't store data itself — it sits on top of HDFS (last week's storage layer) and gives it a
schema and a query language. Four pieces do the work:

- **Driver** — receives the HiveQL statement, manages the session and the query's lifecycle from
  submission to result.
- **Compiler** — parses the query, checks it against the schema the Metastore holds, and turns it
  into an execution plan: historically a MapReduce job, increasingly a Tez or Spark DAG.
- **Metastore** — the piece that makes Hive feel like a database: a separate relational store (often
  its own small MySQL/Derby instance) holding table definitions, column names and types, and which
  HDFS directory each table's (and each partition's) files actually live in. Hive itself never
  touches raw HDFS bytes until a query needs them — schema lookups hit the Metastore, not the data.
- **Execution Engine** — runs the compiled plan against the cluster and streams results back.

```{r hive-metastore}
# A stand-in for what the Metastore actually stores: not the data, just the
# schema and the on-disk location pattern for each table.
metastore <- data.frame(
  table_name = c("orders"),
  columns = c("order_id:int, restaurant_id:int, city:string, order_date:string, delivery_time_minutes:int"),
  partitioned_by = c("order_date"),
  hdfs_location = c("/warehouse/orders/order_date=<date>/")
)
print(metastore, row.names = FALSE)
```

The location pattern matters: `order_date=<date>` in the path is Hive's **partitioning** — each
partition value gets its own HDFS subdirectory, so a query naming a partition in its `WHERE` clause
can skip every directory that isn't that one, without reading a single row from them.

### Managed vs. external tables

```{r managed-vs-external}
table_kinds <- data.frame(
  Kind = c("Managed (internal)", "External"),
  `Hive owns the data?` = c("Yes -- DROP TABLE deletes the HDFS files too",
                             "No -- Hive only owns the schema/metadata"),
  `Typical use` = c("Data Hive itself created or fully controls",
                     "Raw files another pipeline (e.g. Flume, Sqoop) already writes to HDFS"),
  check.names = FALSE
)
print(table_kinds, row.names = FALSE)
```

Getting this choice wrong is a common production mistake: pointing a *managed* table at a shared raw
log directory means one careless `DROP TABLE` in Hive deletes files other pipelines still depend on.
External tables exist specifically so Hive can query data it doesn't own the lifecycle of.

### Partitioning and bucketing: Hive's two data-layout tools

**Partitioning** splits a table's files by a column's value (here, `order_date`) into separate HDFS
directories, so a query that filters on that column reads only the matching directories.
**Bucketing** goes further *within* a partition: it hashes a column into a fixed number of files, so
joins and sampling on that column don't need to scan the whole partition either.

```{r partition-pruning}
# Simulate Hive's partition directories with split() -- one "directory" per
# order_date, exactly matching the metastore's partitioned_by column above.
orders_by_date <- split(orders, orders$order_date)
target_date <- names(orders_by_date)[1]

full_scan_rows <- nrow(orders)
partition_scan_rows <- nrow(orders_by_date[[target_date]])

cat("Query: rows for city='Bangalore' on order_date =", target_date, "\n")
cat("Rows a NAIVE scan of the whole table would read:", full_scan_rows, "\n")
cat("Rows a PARTITION-PRUNED scan actually reads:", partition_scan_rows, "\n")
```

```{r bucketing-demo}
# Bucketing example: hash restaurant_id into 8 buckets within the table.
# A join or sample on restaurant_id only needs to touch the matching bucket
# file, not every partition's full contents.
n_buckets <- 8
orders$bucket <- orders$restaurant_id %% n_buckets
bucket_counts <- table(orders$bucket)
cat("Rows per bucket (should be roughly even across", n_buckets, "buckets):\n")
print(bucket_counts)
```

| Metric | Value | Threshold | Status |
|---|---|---|---|
| Naive full-table scan vs. one partitioned query | ~14x more rows read | should be 1 partition's worth | **Flagged** |
| Bucket row-count spread (max vs. min bucket) | within a few % of even | roughly even across buckets | **Within threshold** |

**Reading the panel:** a query that names `order_date` in its `WHERE` clause and hits a partitioned
table reads roughly a fourteenth of the data a naive scan would — one day's directory instead of all
fourteen. That's the same "don't move what you don't have to" principle as last week's map-local
computation, applied to query planning instead of job scheduling.

### HiveQL in practice

```{r hiveql-example}
# The HiveQL an analyst would actually type -- shown as text since this session
# has no live Hive cluster, with the equivalent result computed in R below so
# you can see the shape of the answer it returns.
hiveql_query <- sprintf("
SELECT city, COUNT(*) AS n_orders, AVG(delivery_time_minutes) AS avg_delivery
FROM orders
WHERE order_date = '%s'
GROUP BY city;
", target_date)
cat(hiveql_query)

# R equivalent, run against the (still-dirty) orders table for one partition
one_day <- orders_by_date[[target_date]]
hiveql_result <- aggregate(delivery_time_minutes ~ city, data = one_day,
                            FUN = function(x) c(n = length(x), avg = round(mean(x, na.rm = TRUE), 1)))
print(hiveql_result)
```

Notice the `NA`-driven fragility here: `AVG()` over a column with missing or negative values gives a
misleading answer (or requires a defensive `WHERE delivery_time_minutes IS NOT NULL` on every query
that touches it). Hive can query dirty data, but it shouldn't have to — which is exactly the job the
next tool exists for.

---

## The data-quality problem Hive alone doesn't solve

```{r dirty-data-check}
n_invalid <- sum(is.na(orders$delivery_time_minutes) | orders$delivery_time_minutes < 0)
pct_invalid <- round(100 * n_invalid / nrow(orders), 1)
cat("Raw order-log rows with a missing or negative delivery_time_minutes:",
    n_invalid, "/", nrow(orders), "(", pct_invalid, "%)\n")
```

| Metric | Value | Threshold | Status |
|---|---|---|---|
| Raw log rows with invalid delivery_time_minutes | ~4% | 0% (every row should be usable) | **Flagged** |

**Reading the panel:** every `AVG(delivery_time_minutes)` query above is already quietly wrong by a
few percent of its input, and a query without the defensive `IS NOT NULL` filter would be far worse —
a `-1` value pulls a city's average down without any error being raised. This is precisely the kind of
raw-log mess an ingestion pipeline (real platforms use Flume or Sqoop, see below) hands to Hadoop
every day, and it needs a cleaning step *before* Hive, not a workaround inside every query written
against it.

---

## Pig: a scripting front end for the cleanup pass

Hive is built for someone who already knows what question they want answered. Pig is built for
someone who needs to describe a multi-step transformation — load, filter, group, join, store — as a
short, readable script, without hand-writing MapReduce for each step. Pig Latin (Pig's language) reads
like a pipeline, not a single query, which is exactly the shape "turn today's raw order log into a
clean table" takes.

```{r piglatin-script, eval=FALSE}
-- The Pig Latin an engineer would actually write for this cleanup job.
-- Not executed here (no live Pig cluster) -- the equivalent runs in R below.
raw_orders  = LOAD 'orders_raw.csv' USING PigStorage(',')
              AS (order_id:int, restaurant_id:int, city:chararray,
                  order_date:chararray, delivery_time_minutes:int);

clean_orders = FILTER raw_orders BY delivery_time_minutes IS NOT NULL
               AND delivery_time_minutes > 0;

by_city      = GROUP clean_orders BY city;

city_stats   = FOREACH by_city GENERATE
               group AS city,
               COUNT(clean_orders) AS n_orders,
               AVG(clean_orders.delivery_time_minutes) AS avg_delivery;

STORE clean_orders INTO '/warehouse/orders_clean' USING PigStorage(',');
```

```{r piglatin-equivalent}
# R equivalent of the Pig Latin above: LOAD -> FILTER -> GROUP -> FOREACH GENERATE.
# `orders` already stands in for the LOAD step (the dataset built in Step 1).
clean_orders <- orders[!is.na(orders$delivery_time_minutes) & orders$delivery_time_minutes > 0, ]

city_stats <- aggregate(delivery_time_minutes ~ city, data = clean_orders,
                         FUN = function(x) c(n = length(x), avg = round(mean(x), 1)))
print(city_stats)

pct_invalid_after <- round(100 * sum(is.na(clean_orders$delivery_time_minutes) |
                                      clean_orders$delivery_time_minutes < 0) / nrow(clean_orders), 1)
cat("\nRows dropped by the FILTER step:", nrow(orders) - nrow(clean_orders),
    "(", round(100 * (nrow(orders) - nrow(clean_orders)) / nrow(orders), 1), "% )\n")
cat("Invalid delivery_time_minutes remaining in clean_orders:", pct_invalid_after, "%\n")
```

| Metric | Value | Threshold | Status |
|---|---|---|---|
| Invalid delivery_time_minutes after the Pig-style FILTER | 0% | 0% | **Within threshold** |

**Reading the panel:** the FILTER step is Pig's equivalent of last week's diagnostic pattern — it
doesn't make the raw data less messy where it landed, it removes the mess before anything downstream
(Hive, in this pipeline) ever queries it. `clean_orders` is what would actually get `STORE`d into the
`/warehouse/orders_clean` directory that a *managed* Hive table would then be pointed at.

```{r hive-on-clean-data}
# Closing the loop: re-run the same HiveQL-shaped query from earlier, now
# against Pig-cleaned data instead of the raw log.
one_day_clean <- clean_orders[clean_orders$order_date == target_date, ]
final_result <- aggregate(delivery_time_minutes ~ city, data = one_day_clean,
                           FUN = function(x) c(n = length(x), avg = round(mean(x), 1)))
print(final_result)
```

---

## Hive vs. Pig: same engine underneath, different job on top

```{r hive-vs-pig}
comparison <- data.frame(
  Aspect = c("Language shape", "Written by", "Best fit"),
  Hive = c("Declarative, SQL-like (HiveQL)", "Analysts who know SQL",
           "\"What's the answer to this question?\""),
  Pig = c("Procedural, step-by-step (Pig Latin)", "Data engineers building pipelines",
          "\"Turn this raw mess into that clean table.\"")
)
print(comparison, row.names = FALSE)
```

Both still compile down to the same batch execution model (MapReduce, or increasingly Tez) that last
week's session covered — the difference is entirely in who's writing the logic and what shape their
problem takes, not in what runs on the cluster underneath.

## The rest of the ecosystem: getting data in and jobs coordinated

Hive and Pig assume the data is already in HDFS and that someone (or something) runs their jobs on a
schedule. Three more tools cover exactly those gaps:

```{r ecosystem-tools}
ecosystem_tools <- data.frame(
  Tool = c("Sqoop", "Flume", "Oozie", "ZooKeeper"),
  Provides = c("Bulk transfer between HDFS and relational databases",
               "Streaming ingestion of continuous log/event data into HDFS",
               "Workflow scheduling that chains jobs (e.g. Sqoop -> Pig -> Hive) together",
               "Distributed coordination: leader election, shared config, locking"),
  Reached_For_When = c("Nightly-import last month's orders table from a production MySQL database",
                        "Continuously stream app/server logs into HDFS as they're generated",
                        "\"Every night at 2am: pull new orders, clean with Pig, then refresh the Hive table\"",
                        "Multiple cluster services need to agree on who's currently in charge")
)
print(ecosystem_tools, row.names = FALSE)
```

Put together, a realistic nightly pipeline reads: **Sqoop/Flume** land raw data in HDFS → **Pig**
cleans and reshapes it → a Hive **managed table** exposes the clean result for querying → **Oozie**
is what actually ran all three steps in order, unattended, every night → **ZooKeeper** keeps the
cluster's own services (YARN's ResourceManager pair, HBase, Kafka where present) agreeing on which
instance is currently active. None of these compete with Hive or Pig — each fills a gap neither one
was built to cover.

---

## Summary

| Check | Result |
|---|---|
| Naive full-table scan vs. partition-pruned query (rows read) | **Flagged** — ~14x more rows without partitioning |
| Bucket row-count spread across 8 buckets | **Within threshold** — roughly even |
| Raw order-log rows with invalid delivery_time_minutes | **Flagged** — ~4% before cleaning |
| Same check after the Pig-style FILTER step | **Within threshold** — 0% |

Hive's value is a schema and a query language over data that was always going to live in HDFS; Pig's
value is a readable pipeline for the transformation that has to happen before that data is trustworthy
enough to query. Neither replaces MapReduce — both are what most people actually touch instead of it.
